A simple visual alternative to all pairwise comparisons

### 2023-05-19 text/22/analysis-of-means.Rmd
Added ai-fairness.

---
title: "A second alternative to the pairwise testing approach"
author: "Steve Simon"
date: '2022-07-04'
output:
  html_document: default
category: Blog post
tags: Analysis of means
source: new
---

In an earlier blog post, I mentioned Dunnett's test, a comparison of multiple treatments to a control group as an alternative to the pairwise approach to testing that is commonly used in analysis of variance. Here's a second interesting alternative.

### Comparison to an overall mean

You are in a setting where you are expecting to see a degree of uniformity among several groups. For example, you have a group of clinics who all treat the same type of patients. The time spent with each patient, the total cost of the medications prescribed, the length of the notes written in the medical record--these can differ from patient to patient. But because of how the patients are assigned to a particular clinic (more or less randomly), these should balance out on average.

If there is one clinic that has different average results, it has to be because that the physicians at that clinic have a different approach to treatment. If you can identify a clinic (it could be more than one clinic, of course) that differs from the norm, then investigate. Perhaps that clinic is doing something "wrong" that causes the discrepancy, or maybe that clinic is the only one doing things right. In any case understanding why you see unexpected deviations in a process that should be uniform across clinics provides you with an opportunity to learn something.

@@

This approach is not that common, but you can find it in most statistical software programs under the name "Analysis of Means." It fits in well with a setting where you hope that all the groups produce consistent results. If any do not, then you want to identify the group or groups that deviate from the norm and study factors that may account for the deviation.

The analysis of means approach compares each group mean to the overall mean. It is easy to implement and it lends itself to a simple graphical display. You need a table of critical values, which depend on alpha (the overall Type I error rate), t (the number of groups), and n (the number of observations within each group). Some tables use the degrees of freedom for error in place of n.

@@

Five genetically modified strains of eggplant were tested for presence of a natural insecticide, CryA1c, in the leaves of the plant. Four plants were assessed for each strain. Here are the descriptive statistics. The research question is whether any strain differs from the overall mean level of CryA1c.

@@

@@

https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0157498

@@

The formulas get a bit more complex and the graph gets a bit messier if the sample size differs from one group to another.

More importantly, the only comparisons that you can make with the analysis of means approach is a comparison to the overall mean. If two groups are both statistically significantly higher than the overall mean, you cannnot make a comparison between those two groups without losing control over the overall Type I error rate.

@@

The Analysis of Means, just like the Dunnett's test described in an earlier blog post, provides a simple approach to testing a different type of hypothesis. Because it reduces the number of comparisons from all possible pairwise differences to a comparison of each group to the overall mean, you gain some precision and can summarize your results in a simple easily understood graph.

>>>>

---
title: "A simple visual alternative to all pairwise comparisons"
author: "Steve Simon"
date: '2022-07-04'
output:
  html_document: default
category: Blog post
tags: Analysis of means
source: new
---

The classic follow-up to a significant F test in analysis of variance is examining all pairwise differences in means. It's great for many problems, but sometimes you may not want to focus on all pairwise differences. Sometimes, there are so many pairwise differences that you get lost in all the clutter. Other times, you are trying to find out if, like the Sesame Street song, one of these groups is not like the others. Perhaps, you are trying to assure that groups conform to a common standard. Analysis of Means (ANOM) helps in all these settings.

@@

This approach is not that common, but you can find it in most statistical software programs under the name "Analysis of Means" or its acronym, ANOM. It fits in well with a setting where you hope that all the groups produce consistent results. If any do not, then you want to identify the group or groups that deviate from the norm and study factors that may account for the deviation.

The ANOM approach compares each group mean to the overall mean. It is easy to implement and it lends itself to a simple graphical display. You need a table of critical values, which depend on alpha (the overall Type I error rate), t (the number of groups), and n (the number of observations within each group). Some tables use the degrees of freedom for error in place of n.

@@

Imagine that you are a crop scientist, comparing five genetically modified strains of eggplant. You developed these strains to produce a natural insecticide, CryA1c, in the eggplant leaves.

You hope that all strains produce a comparable amount of insecticide. If any strain produces a lot more or a lot less insecticide, you will investigate full genetic profile of that strain to see why it is different from the others.

@@

The data comes from

@@

<!--- https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0157498 --->

and my description of the testing scenario is a bit simpler than the analysis done in that paper.

@@

The graph gets a bit messier if the sample size differs from one group to another.

More importantly, the only comparisons that you can make with the analysis of means approach is a comparison to the overall mean. If two groups are both statistically significantly higher than the overall mean, you cannnot make a comparison between those two groups without losing control of the overall Type I error rate.

@@

The Analysis of Means provides a simple approach to testing a different type of hypothesis. Because it reduces the number of comparisons from all possible pairwise differences to a comparison of each group to the overall mean, you gain some precision and can summarize your results in a simple easily understood graph.
